Equalizer\\Programming and User Guide

### 2012-06-29 Developer/ProgrammingGuide/paper.tex
1.4 additions

\date{
  \vspace{14cm}
  \begin{textblock}{}(0,4.5)
    \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{textblock}
      {\Large The official reference for developing and
        deploying parallel, scalable OpenGL\texttrademark\ applications using
        the Equalizer parallel rendering framework}\\\vspace{1cm}
      Version 1.11.3 for Equalizer 1.4-beta\\[\medskipamount]
  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
    1.10    & Feb 17, 2012 & Revision for Equalizer 1.2\\
    1.11.1  & Apr 10, 2012 & Minor revision\\
    1.11.2  & Jun 25, 2012 & Add \sref{sScaling} and \sref{sROI}\\
    1.11.3  & Jun 26, 2012 & Add \sref{sZeroconf}, update \sref{sNetObject2} and \aref{aFileFormat}\\
    \hline
  \end{tabularx}}
\clearpage

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eqHello.png}
  {\caption{\label{fHello}Hello, World!}}
\end{wrapfigure}
The \textsf{eqHello} example is a minimal application to illustrate the basic
principle of any Equalizer application: The application developer has to
implement a rendering method similar to the \textsf{glutDisplayFunc} in GLUT
applications.

It can be run as a stand-alone application from the command line or using an
explicit configuration file for a visualization cluster. In the stand-alone
case, any Equalizer application including eqHello, will automatically
configure itself to use all graphics cards found on the local system for
scalability.

\subsection{Sequel}

Sequel is a thin layer on top of the canonical Equalizer programming
interface. It substantially simplifies the creation of parallel rendering
applications by implementing proven design patterns of Equalizer
applications. Sequel does not prohibit the usage of Equalizer or Collage
functionality, it simply makes it easier to use for the common use cases.

The main function instantiates an application object, initializes it, starts the
main loop and finally de-initializes the application:

{\footnotesize\lstinputlisting[firstline=58, lastline=66]
  {../../../Equalizer/examples/eqHello/hello.cpp}}

The application is responsible for creating renderers. Sequel will request one
renderer for each GPU rendering thread. Sequel also provides automatic mapping
and synchronization of distributed objects, for which the application has to
provide a creation callback. Since \textsf{eqHello} does not need any
distributed objects it provides an empty implementation of
\textsf{createObject}:

{\footnotesize\lstinputlisting[firstline=47, lastline=55]
  {../../../Equalizer/examples/eqHello/hello.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=69, lastline=81]
  {../../../Equalizer/examples/eqHello/hello.cpp}}

@@

The API documentation on the website or in the header files provides a
comprehensive documentation on individual methods, types and other elements of
the API. Methods marked with a specific version are part of the official, public
API and have been introduced by this Equalizer version. Reasonable care is taken
to not break API compatibility or change the semantics of these methods within
future Equalizer versions of the same major revision. Any eventual changes to
the public API is documented in the release notes and the file
\textsf{CHANGES.txt}.

@@

Application developers are exposed to the following namespaces:

\begin{description}
\item[eq] The core namespace for the Equalizer client library. The
  classes and their relationship in this namespace closely model the
  configuration file format. The classes in the \textsf{eq} namespace
  are the main subject of this Programming Guide.
\item[eq::util] The eq::util namespace provides common utility classes, which
  often simplify the usage of OpenGL functions. Most classes in this namespace
  are used by the Equalizer client library, but are usable independently
  from Equalizer for application development.
\item[eq::admin] The eq::admin namespace implements an administrative API to
  change the configurations of a running server. This admin API is not yet
  finalized and will very likely change in the future.
\item[eq::fabric] The eq::fabric namespace is the common data management and
  transport layer between the client library, the server and the administrative
  API. Most Equalizer client classes inherit from base classes in this namespace
  and have all their data access methods in these base classes.
\item[co] Collage is the network library used by Equalizer. It provides basic
  functionality for network communication, such as \textsf{Connection} and
  \textsf{ConnectionSet}, as well as higher-level functionality such as
  \textsf{Node}, \textsf{LocalNode}, \textsf{Object} and
  \textsf{Serializable}. Please refer to \sref{sNetwork} for an introduction
  into the network layer, and to \sref{sNetObject} for distributed objects.
\item[lunchbox] The lunchbox library provides C++ classes to abstract the
  underlying operating system and to implement common helper functionality for
  multi-threaded applications. Examples are \textsf{lunchbox::Clock} providing a
  high-resolution timer, or \textsf{lunchbox::MTQueue} providing a thread-safe,
  blocking fifo. Classes in this namespace are fully documented in the API
  documentation on the Equalizer website, and are not subject of this
  Programming Guide.
\end{description}

Examples shipped with Equalizer are implemented in their own namespaces, e.g.,
\textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing from the
\textsf{eq} namespace, with the occasional usage of functionality from the
\textsf{eq::util}, \textsf{eq::fabric}, \textsf{co} and \textsf{lunchbox}
namespace.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/namespaces}
  {\caption{\label{fNamespaces}Namespace Layering for eqPly and eqServer}}
\end{wrapfigure}
The functionality of the Equalizer server is implemented in the
\textsf{eq::server} namespace, which is an internal namespace not to be used by
application developers. The \textsf{eq::admin} namespace enables run-time
administration of Equalizer servers. These namespaces do not yet expose a stable
API. \fig{fClientUml} provides an overview map of the most important classes in
the Equalizer namespace, grouped by functionality.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\label{fModel}Simplified Execution Model}}
\end{wrapfigure}
The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

The rendering threads concurrently render the application's
database. The data\-base should be accessed in a read-only fashion
during rendering to avoid threading problems. This is normally the case,
for example all modern scene graphs use read-only render traversals, writing the
GPU-specific information into a separate per-GPU data structure.

@@

\section{\label{sEqPly}The eqPly Polygonal Renderer}

In this section the source code of \textsf{eqPly} is explained in
detail, and relevant design decisions, caveats and other remarks are
discussed.

\textsf{eqPly} is a parallel renderer for polygonal data in the
\textsf{ply} file format. It supports all relevant Equalizer features,
and can be used to render on large-scale displays, immersive
environments with head tracking and to render massive data sets using
all scalable rendering features of Equalizer.

@@

%\begin{wrapfigure}{r}{.618\textwidth}
\begin{figure}[ht!]\center
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\label{fObjects}Scene Graph Distribution}}
\end{figure}
Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

@@

\subsubsection{Rendering}

@@

The model data is spatially organized in a 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree. During
this preprocessing step, each node of the tree gets a database range
assigned. The root node has the range [0, 1], its left child [0, 0.5]
and its right child [0.5, 1], and so on for all nodes in the tree. The
preprocessed model is saved in a binary format for accelerating
subsequent loading.

\begin{wrapfigure}{r}{.618\textwidth}
%\begin{figure}[h!t]\center
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\label{fRender}Main Render Loop}}
\end{wrapfigure}
%\end{figure}
The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visibility against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list.

@@

Events are received by an event handler. The event handler finds the
\textsf{eq::Sys\-tem\-Win\-dow} associated to the event. It then creates a
generic \textsf{Event}, which holds the event data in an independent format. The
original native event and this generic \textsf{Event} form the
\textsf{SystemWindowEvent}, which is passed to the concrete
\textsf{SystemWindow} for processing.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
The purpose of the \textsf{SystemWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further, generic processing, the \textsf{Event} is passed on to
\textsf{Window::processEvent}. This \textsf{Event} no longer contains
the native event.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\label{fEventUML}UML Class Diagram for Event Handling}}
\end{wrapfigure}
For each GLX window, one \textsf{GLXEventHandler} is allocated. FBO's and
PBuffers are handled in the same way as window drawables. The event dispatch
finds the corresponding event handler for each received event, which is then
used to process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::\-process\-Event}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\label{fThreads}Threads within one Node Process}}
\end{wrapfigure}
During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

@@

The receiver and command thread are terminated when the application
stops network communications using \textsf{Client::exitLocal}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \subfigure[]{\label{fROIDB2Way}
    \includegraphics[width=.29\textwidth]{images/roiDB2way.png}
  }\hfil
  \subfigure[]{\label{fROIDB4Way}
    \includegraphics[width=.29\textwidth]{images/roiDB4way.png}
  }%
  {\caption{\label{fROIDB}ROI for a two-way (a) and four-way DB (b) compound}}
\end{wrapfigure}
The first optimization concerns the compositing operation. The declared ROI
allows Equalizer to restrict the readback, compression, network transmission and
assembly to only the relevant pixels.

This is particularly important for DB compounds, which typically cover less
screen space on each channel as resources are added to the
decomposition. \fig{fROIDB} illustrates this for a two-way and four-way
decomposition. Declaring the ROI alleviates the increasing compositing cost as
resources are added to a DB compound.

@@

The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\label{fDirectSend}Direct Send Compositing}}
\end{wrapfigure}
In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

@@

%\begin{wrapfigure}{r}{.618\textwidth}
\begin{figure}[h!t]\center
  \includegraphics[width=.9\textwidth]{images/compositor.pdf}
  {\caption{\label{fCompositor}Functional Diagram of the Compositor}}
\end{figure}
On a high level, the compositor combines multiple input frames using 2D
tiling, depth-compositing for polygonal data or sorted, alpha-blended
compositing for semitransparent volumetric data. These operations
composite either directly all images on the GPU, or use a CPU-based
compositor and then transfer the preintegrated result to the GPU. The
high-level entry points automatically select the best algorithm. The
CPU-based compositor uses, based on availability, OpenMP and the
Paracomp compositing library to accelerate its operation.

@@

When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}
\end{wrapfigure}
Conceptually, the individual volume bricks of each of the source
channels produces pixel data which can be handled like one big 'slice'
through the full texture. Therefore they have to be blen\-ded
back-to-front in the same way as the slice planes are blended during
rendering.

@@

\begin{wrapfigure}{l}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
For recomposition, the 2D frame buffer contents are blended to form a
seamless picture. For correct blending, the frames are ordered in the
same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
\end{wrapfigure}
The master instance can \textsf{sync} delta data received from slaves. When
using the version generated by the slave object commit, only the delta for this
version is applied. When using \textsf{VERSION\_NEXT} as the version parameter
to \textsf{sync}, one slave commit is unpacked. This call may block if no data
is pending. When using \textsf{VERSION\_HEAD} all pending versions are
unpacked. The data is synchronized in the order it has been received from the
slaves. If a specific order is needed, it is the application's responsibility to
serialize the slave commits accordingly.

>>>>

\date{
  \vspace{14cm}
  \begin{textblock}{}(0,4.5)
    \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{textblock}
      {\Large The official reference for developing and
        deploying parallel, scalable OpenGL\texttrademark\ applications using
        the Equalizer parallel rendering framework}\\\vspace{1cm}
      Version 1.11.4 for Equalizer 1.4 beta\\[\medskipamount]
  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
    1.10    & Feb 17, 2012 & Revision for Equalizer 1.2\\
    1.11.1  & Apr 10, 2012 & Minor revision\\
    1.11.2  & Jun 25, 2012 & Add \sref{sScaling} and \sref{sROI}\\
    1.11.3  & Jun 26, 2012 & Add \sref{sZeroconf}, update \sref{sNetObject2} and \aref{aFileFormat}\\
    1.11.4  & Jun 29, 2012 & Add \sref{sSeqPly}\\
    \hline
  \end{tabularx}}
\clearpage

@@

The \textsf{eqHello} example is a minimal application to illustrate the basic
principle of any Equalizer application: The application developer has to
implement a rendering method similar to the \textsf{glutDisplayFunc} in GLUT
applications.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eqHello.png}
  {\caption{\label{fHello}Hello, World!}}
\end{wrapfigure}
It can be run as a stand-alone application from the command line or using an
explicit configuration file for a visualization cluster. In the stand-alone
case, any Equalizer application, including eqHello, will automatically configure
itself to use all graphics cards found on the local system for scalability.

The \textsf{eqHello} example uses Sequel, a thin layer on top of the canonical
Equalizer programming interface. \sref{sSeqPly} introduces Sequel in detail, and
\sref{sEqPly} introduces the full scope of the Equalizer API.

The main \textsf{eqHello} function instantiates an application object,
initializes it, starts the main loop and finally de-initializes the application:

{\footnotesize\lstinputlisting[firstline=56, lastline=64]
  {../../../Equalizer/examples/eqHello/hello.cpp}}

The application object represents one process in the cluster. The primary
application instance has the rendering loop and controls all execution. All
other instances used for render client processes are passive and driven by
Equalizer. The application is responsible for creating the renderers, of which
one per GPU is used:

{\footnotesize\lstinputlisting[firstline=47, lastline=53]
  {../../../Equalizer/examples/eqHello/hello.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=67, lastline=79]
  {../../../Equalizer/examples/eqHello/hello.cpp}}

@@

The API documentation is available on the website or in the header files, and
provides a comprehensive documentation on individual methods, types and other
elements of the API. Methods marked with a specific version are part of the
official, public API and have been introduced by this Equalizer
version. Reasonable care is taken to not break API compatibility or change the
semantics of these methods within future Equalizer versions of the same major
revision. Any eventual changes to the public API are documented in the release
notes and the file \textsf{CHANGES.txt}.

@@

The Equalizer software stack is modularized, layering gradually more powerful,
but less generic APIs on to of each other. It furthermore relies on a number of
required and optional libraries. Application developers are exposed to the
following name\-spa\-ces:

\begin{description}
\item[seq] The core namespace for Sequel, the simple interface to the Equalizer
  client library.
\item[eq] The core namespace for the Equalizer client library. The classes and
  their relationship in this namespace closely model the configuration file
  format. The classes in the \textsf{eq} namespace are the main subject of this
  Programming Guide.\fig{fClientUml} provides an overview map of the most
  important classes in the Equalizer namespace, grouped by functionality.
\item[eq::util] The eq::util namespace provides common utility classes, which
  often simplify the usage of OpenGL functions. Most classes in this namespace
  are used by the Equalizer client library, but are usable independently
  from Equalizer for application development.
\item[eq::admin] The eq::admin namespace implements an administrative API to
  change the configurations of a running server. This admin API is not yet
  finalized and will very likely change in the future.
\item[eq::fabric] The eq::fabric namespace is the common data management and
  transport layer between the client library, the server and the administrative
  API. Most Equalizer client classes inherit from base classes in this namespace
  and have all their data access methods in these base classes.
\item[co] Collage is the network library used by Equalizer. It provides basic
  functionality for network communication, such as \textsf{Connection} and
  \textsf{ConnectionSet}, as well as higher-level functionality such as
  \textsf{Node}, \textsf{LocalNode}, \textsf{Object} and
  \textsf{Serializable}. Please refer to \sref{sNetwork} for an introduction
  into the network layer, and to \sref{sNetObject} for distributed objects.
\item[lunchbox] The lunchbox library provides C++ classes to abstract the
  underlying operating system and to implement common helper functionality for
  multi-threaded applications. Examples are \textsf{lunchbox::Clock} providing a
  high-resolution timer, or \textsf{lunchbox::MTQueue} providing a thread-safe,
  blocking fifo. Classes in this namespace are fully documented in the API
  documentation on the Equalizer website, and are not subject of this
  Programming Guide.
\item[hwloc, boost, vmmlib, gpu-sd] External libraries providing manual and
  automatic thread affinity, serialization and the foundation for RSP multicast,
  vector and matrix mathematics as well as local and remote GPU discovery,
  respectively. The hwloc and gpu-sd libraries are used only internally and not
  exposed through the API.
\item[eq::server] The server namespace, implementing the functionality of the
  Equalizer server, which is an internal namespace not to be used by application
  developers. The \textsf{eq::admin} namespace enables run-time administration
  of Equalizer servers. The server does not yet expose a stable
  API.
\end{description}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/namespaces}
  {\caption{\label{fNamespaces}Namespaces}}
\end{wrapfigure}
The Equalizer examples are implemented in their own namespaces, e.g.,
\textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing from the
\textsf{eq} namespace, with the occasional usage of functionality from the
\textsf{eq::util}, \textsf{eq::fabric}, \textsf{co} and \textsf{lunchbox}
namespace. \fig{fNamespaces} shows the namespaces and their layering.

@@

The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\label{fModel}Simplified Execution Model}}
\end{wrapfigure}
The rendering threads concurrently render the application's
database. The data\-base should be accessed in a read-only fashion
during rendering to avoid threading problems. This is normally the case,
for example all modern scene graphs use read-only render traversals, writing the
GPU-specific information into a separate per-GPU data structure.

@@

\section{\label{sSeqPly}Sequel: The seqPly Polygonal Renderer}

In this section the source code of \textsf{seqPly} is used to introduce the
Sequel API in detail, and relevant design decisions, caveats and other remarks
are discussed. Sequel is conceived to lower the entry barrier into creating
parallel rendering applications without sacrificing major functionality. It
implements proven design patterns of Equalizer applications. Sequel does not
prohibit the usage of Equalizer or Collage functionality, it simply makes it
easier to use for the common use cases.

The \textsf{seqPly} parallel renderer provides a subset of features found in its
cousin application, \textsf{eqPly}, introduced in the next section. It is more
approachable, using approximately one tenth of the source code needed by
\textsf{eqPly} due to the use of the Sequel abstraction layer and reduced
functionality.

The main function is almost identical to \textsf{eqHello}. It instantiates an
application object, initializes it, starts the main loop and finally
de-initializes the application. The source code is not reproduced here due to
its similarity with \sref{sHello}.

Sequel applications derive their application object from
\textsf{seq::Application} and selectively override functionality. They have to
implement \textsf{createRenderer}, as explained below. The \textsf{seqPly}
application overrides init, exit, run and implements createRenderer.

The initialization and exit routines are overwritten to parse
\textsf{seqPly}-specific command line options and to load and unload the
requested model:

{\footnotesize\lstinputlisting[firstline=41, lastline=49]
  {../../../Equalizer/examples/seqPly/application.cpp}}
{\footnotesize\lstinputlisting[firstline=56, lastline=60]
  {../../../Equalizer/examples/seqPly/application.cpp}}

Sequel manages distributed objects, simplifying their use. This includes
registration, automatic creation and mapping as well as commit and
synchronization of objects. One special object for initialization (not used in
\textsf{seqPly}) and one for per-frame data are managed by Sequel, in addition
to an arbitrary number of application-specific objects.

The objects passed to \textsf{seq::Application::init} and
\textsf{seq::Application::run} are automatically distributed and instantiated on
the render clients, and then passed to the respective task callback methods. The
application may pass a 0 pointer if it does not need an object for
initialization or per-frame data. Objects are registered with a type, and when
automatically created, the \textsf{createObject} method on the application or
renderer is used to create an instance based on this type.

The run method is overloaded to pass the frame data object to the Sequel
application run loop. The object will be distributed and synchronized to all
renderers:

{\footnotesize\lstinputlisting[firstline=51, lastline=55]
  {../../../Equalizer/examples/seqPly/application.cpp}}

The application is responsible for creating renderers. Sequel will request one
renderer for each GPU rendering thread. Sequel also provides automatic mapping
and synchronization of distributed objects, for which the application has to
provide a creation callback:

{\footnotesize\lstinputlisting[firstline=62, lastline=77]
  {../../../Equalizer/examples/seqPly/application.cpp}}

The \textsf{seqPly} rendering code uses the same data structure and algorithm as
\textsf{eqPly}, described in \sref{sRendering}. This renderer captures the
GPU-specific data in a \textsf{State} object, which is created and destroyed
during init and exit. The state captures also the OpenGL function table, which
is available when init is called, but not yet during the constructor of the
renderer:

{\footnotesize\lstinputlisting[firstline=46, lastline=59]
  {../../../Equalizer/examples/seqPly/renderer.cpp}}

The rendering code is similar to the typical OpenGL rendering code, except for a
few modifications to configure the rendering. First, the render context is
applied and lighting is set up. The render context, described in detail in
\sref{sRenderContext}, sets up the stereo buffer, 3D viewport as well as the
projection and view matrices using the appropriate OpenGL calls. Applications
can also retrieve the render context and apply the settings themselves:

{\footnotesize\lstinputlisting[firstline=69, lastline=80]
  {../../../Equalizer/examples/seqPly/renderer.cpp}}

After the static light setup, the model matrix is applied to the existing view
matrix, completing the modelview matrix and positioning the model. Sequel
maintains a per-view camera, which is modified through mouse and keyboard events
and determines the model matrix. Applications can overwrite this event handling
and maintain their own camera model. Afterwards, the state is set up with the
projection-modelview matrix for view frustum culling, the DB range for sort-last
rendering and the cull-draw traversal is executed, as described in
\sref{sRendering}:

{\footnotesize\lstinputlisting[firstline=81, lastline=97]
  {../../../Equalizer/examples/seqPly/renderer.cpp}}

\section{\label{sEqPly}Equalizer: The eqPly Polygonal Renderer}

In this section the source code of \textsf{eqPly} is used to introduce the
Equalizer API in detail, and relevant design decisions, caveats and other
remarks are discussed.

\textsf{eqPly} is a parallel renderer for polygonal data in the
\textsf{ply} file format. It supports all relevant Equalizer features,
and can be used to render on large-scale displays, immersive
environments with head tracking and to render massive data sets using
all scalable rendering features of Equalizer. It is a superset of
\textsf{seqPly}, introduced in \sref{sSeqPly}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
%\begin{figure}[ht!]\center
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\label{fObjects}Scene Graph Distribution}}
%\end{figure}
\end{wrapfigure}
Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

@@

\subsubsection{\label{sRendering}Rendering}

@@

\begin{wrapfigure}{r}{.618\textwidth}
%\begin{figure}[h!t]\center
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\label{fRender}Main Render Loop}}
\end{wrapfigure}
%\end{figure}
The model data is spatially organized in a 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree. During
this preprocessing step, each node of the tree gets a database range
assigned. The root node has the range [0, 1], its left child [0, 0.5]
and its right child [0.5, 1], and so on for all nodes in the tree. The
preprocessed model is saved in a binary format for accelerating
subsequent loading.

The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visibility against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\label{fEventProcessing}Event Processing}}\vspace{-4ex}
\end{wrapfigure}
Events are received by an event handler. The event handler finds the
\textsf{eq::Sys\-tem\-Win\-dow} associated to the event. It then creates a
generic \textsf{Event}, which holds the event data in an independent format. The
original native event and this generic \textsf{Event} form the
\textsf{SystemWindowEvent}, which is passed to the concrete
\textsf{SystemWindow} for processing.

The purpose of the \textsf{SystemWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further, generic processing, the \textsf{Event} is passed on to
\textsf{Window::processEvent}. This \textsf{Event} no longer contains
the native event.

@@

For each GLX window, one \textsf{GLXEventHandler} is allocated. FBO's and
PBuffers are handled in the same way as window drawables. The event dispatch
finds the corresponding event handler for each received event, which is then
used to process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::\-process\-Event}.

@@

During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\label{fThreads}Threads within one Node Process}}
\end{wrapfigure}
The receiver and command thread are terminated when the application
stops network communications using \textsf{Client::exitLocal}.

@@

The first optimization concerns the compositing operation. The declared ROI
allows Equalizer to restrict the readback, compression, network transmission and
assembly to only the relevant pixels.

\begin{wrapfigure}{l}{.618\textwidth}
  \subfigure[]{\label{fROIDB2Way}
    \includegraphics[width=.29\textwidth]{images/roiDB2way.png}
  }\hfil
  \subfigure[]{\label{fROIDB4Way}
    \includegraphics[width=.29\textwidth]{images/roiDB4way.png}
  }%
  {\caption{\label{fROIDB}ROI for a two-way (a) and four-way DB (b) compound}}
\end{wrapfigure}
This is particularly important for DB compounds, which typically cover less
screen space on each channel as resources are added to the
decomposition. \fig{fROIDB} illustrates this for a two-way and four-way
decomposition. Declaring the ROI alleviates the increasing compositing cost as
resources are added to a DB compound.

@@

Direct send has two important properties: an algorithmic complexity of O(1) for
each node, that is, the compositing cost per node is constant as resources are
added, and the capability to perform total ordering during compositing, e.g, to
back-to-front sort all contributions of a 3D volume rendering correctly.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\label{fDirectSend}Direct Send Compositing}}
\end{wrapfigure}
The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

%\begin{wrapfigure}{r}{.618\textwidth}
\begin{figure}[h!t]\center
  \includegraphics[width=.9\textwidth]{images/compositor.pdf}
  {\caption{\label{fCompositor}Functional Diagram of the Compositor}}
\end{figure}

@@

On a high level, the compositor combines multiple input frames using 2D
tiling, depth-compositing for polygonal data or sorted, alpha-blended
compositing for semitransparent volumetric data. These operations
composite either directly all images on the GPU, or use a CPU-based
compositor and then transfer the preintegrated result to the GPU. The
high-level entry points automatically select the best algorithm. The
CPU-based compositor uses, based on availability, OpenMP and the
Paracomp compositing library to accelerate its operation.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\label{fSlices}Volume Rendering using 3D Textures}}\vspace{-2ex}
\end{wrapfigure}
When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

Conceptually, the individual volume bricks of each of the source
channels produces pixel data which can be handled like one big 'slice'
through the full texture. Therefore they have to be blen\-ded
back-to-front in the same way as the slice planes are blended during
rendering.

@@

For recomposition, the 2D frame buffer contents are blended to form a
seamless picture. For correct blending, the frames are ordered in the
same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

@@

The master instance can \textsf{sync} delta data received from slaves. When
using the version generated by the slave object commit, only the delta for this
version is applied. When using \textsf{VERSION\_NEXT} as the version parameter
to \textsf{sync}, one slave commit is unpacked. This call may block if no data
is pending. When using \textsf{VERSION\_HEAD} all pending versions are
unpacked. The data is synchronized in the order it has been received from the
slaves. If a specific order is needed, it is the application's responsibility to
serialize the slave commits accordingly.
